\singlespacing \vspace{-1.6cm} \LARGE Covariation among reproductive traits in flowering plants shapes interactions with floral visitors

### 2022-10-04 Manuscript/Lanuza_et_al_2022.tex
edits

\setstretch{2}
\vspace{-1.4cm}

\textbf{Jose B. Lanuza$^{1,2}$} \orcidlink{0000-0002-0287-409X}, \textbf{Romina Rader$^{1}$} \orcidlink{0000-0001-9056-9118}, \textbf{Jamie Stavert$^{3}$} \orcidlink{0000-0002-2103-5320}, \textbf{Liam K. Kendall$^{4}$} \orcidlink{0000-0002-0671-0121}, \textbf{Manu E. Saunders$^{1}$} \orcidlink{0000-0003-0645-8277} \textbf{and Ignasi Bartomeus$^{2}$} \orcidlink{0000-0001-7893-4389}

\small

\textsuperscript{1} School of Environmental and Rural Science, University of New England, Armidale, New South Wales 2350, Australia. \textsuperscript{2} Estación Biológica de Doñana (EBD-CSIC), E-41092 Seville, Spain. \textsuperscript{3} Department of Conservation \textbar{} Te Papa Atawhai, Auckland, New Zealand. \textsuperscript{4} Centre for Environmental and Climate Science, Lund University, Sölvegatan 37, S-223 62 Lund, Sweden.

\doublespacing
\normalsize

\textbf{Author for correspondence:} Jose B. Lanuza \textbar{} \href{mailto:barragansljose@gmail.com}{\nolinkurl{barragansljose@gmail.com}}

@@

Flowering plants display an astonishing diversity of reproductive structures {[}\protect\hyperlink{ref-barrett2002}{1},\protect\hyperlink{ref-schiestl2013}{2}{]} that shape their interactions with animal pollinators {[}\protect\hyperlink{ref-fenster2004}{3},\protect\hyperlink{ref-dellinger2020}{4}{]}. The variety of reproductive strategies used by different plant species, from specialized reliance on a single pollinator, to full self-pollination, is incredibly rich. However, not all reproductive trait combinations in plants are possible due to evolutionary and ecological constraints {[}\protect\hyperlink{ref-stearns1989}{5},\protect\hyperlink{ref-agrawal2020}{6}{]}. Despite recent advances in theoretical and empirical understanding of correlations between plant reproductive traits at large ecological scales {[}\protect\hyperlink{ref-salguero2016}{7}--\protect\hyperlink{ref-song2022}{11}{]}, studies typically focus on just a few traits, often in isolation. Thus, to develop a comprehensive understanding of the plant reproductive spectrum of trait variation, as done recently for root {[}\protect\hyperlink{ref-laughlin2021}{12}{]}, leaf {[}\protect\hyperlink{ref-wright2004}{13}{]} and wood {[}\protect\hyperlink{ref-chave2009}{14}{]} related traits, we require a multi-trait perspective with broad geographical coverage. Importantly, characterization of the plant reproductive spectrum could reveal key reproductive strategies among different species {[}\protect\hyperlink{ref-agrawal2020}{6}{]} and improve knowledge of associations between specific reproductive traits and pollinators {[}\protect\hyperlink{ref-roddy2021}{10}{]}.

With the recent availability of large trait databases {[}e.g., TRY \protect\hyperlink{ref-kattge2011}{15},COMPADRE \protect\hyperlink{ref-salguero2015}{16}{]}, there has been increased research on plant ecological strategies, which has produced identification of global patterns in plant form and function {[}\protect\hyperlink{ref-salguero2016}{7},\protect\hyperlink{ref-diaz2016}{17}--\protect\hyperlink{ref-carmona2021}{19}{]}. However, studies that look at multiple traits tend to overlook plant reproductive traits {[}\protect\hyperlink{ref-roddy2021}{10},\protect\hyperlink{ref-evojtko2020}{20},\protect\hyperlink{ref-ruger2018}{21}{]}, and focus on trait correlations in other plant organs such as morphological and physiological leaf traits {[}\protect\hyperlink{ref-wright2004}{13},\protect\hyperlink{ref-shipley2006}{22}--\protect\hyperlink{ref-osnas2013}{24}{]}. Despite the lack of a holistic understanding that depicts reproductive trait covariation patterns, there are widely recognized associations between plant reproductive trait pairs (e.g., the negative correlation between flower size and flower number {[}\protect\hyperlink{ref-sargent2007}{25}--\protect\hyperlink{ref-vojtko2022}{27}{]}; the positive association between flower size and outcrossing rate {[}\protect\hyperlink{ref-goodwillie2010}{28}{]}; or the positive correlation between outcrossing rate and lifespan {[}\protect\hyperlink{ref-barrett2003}{29}--\protect\hyperlink{ref-moeller2017}{31}{]}). Although identification of these trait correlations has allowed progress toward a conceptual framework that integrates different floral trait relationships {[}\protect\hyperlink{ref-roddy2021}{10}{]}, we still lack empirical understanding of how these associations among floral and other reproductive traits (e.g., reproductive biology) shape plant life history strategies.

Trait-based approaches {[}\protect\hyperlink{ref-fenster2004}{3},\protect\hyperlink{ref-rosas2014}{32}{]} and trait-matching analyses {[}\protect\hyperlink{ref-stang2009}{33},\protect\hyperlink{ref-bartomeus2016}{34}{]} are powerful for exploring the underlying mechanisms driving plant-pollinator interactions. For example, plant traits can define species' network roles {[}e.g., specialists vs generalists \protect\hyperlink{ref-tur2013}{35}{]} and plant species that occupy reproductive trait space extremes are likely to be more specialized and display tighter trait-matching with pollinators {[}\protect\hyperlink{ref-junker2013}{36},\protect\hyperlink{ref-coux2016}{37}{]}. Indeed, the degree of morphological matching between plants and floral visitors often determines whether or not interactions occur, and consequently influences the broader network structure {[}\protect\hyperlink{ref-stang2009}{33},\protect\hyperlink{ref-ibanez2012}{38}{]}. However, it is still unclear how specific reproductive traits (e.g., mating or compatibility system) influence plant-pollinator associations {[}\protect\hyperlink{ref-tur2013}{35},\protect\hyperlink{ref-sargent2008}{39},\protect\hyperlink{ref-devaux2014}{40}{]}. Indeed, both morphology and mating system can determine a species' functional role within pollination networks, and the combination of several traits increases power for predicting network structure {[}\protect\hyperlink{ref-eklof2013}{41}{]}. Because the species position within the multidimensional trait space can result in different interaction patterns or adaptations to pollinators {[}see \protect\hyperlink{ref-dehling2016}{42} for a purely morphological approach{]}, the use of the reproductive trait space to explore plant-pollinator interactions could help to identify species functional roles and associations of traits with specific pollinators (i.e., pollination syndromes).

@@

We selected 28 studies from 18 different countries that constituted a total of 64 plant-pollinator networks (see table S1 and figure S1). These studies recorded plant-pollinator interactions in natural systems and were selected so that we had broad geographical representation across different ecological communities. Although these studies differ in sampling effort and methodology, all studies provided information about plant-pollinator interactions (weighted and non-weighted), which we used to build a database of plant species that are likely to benefit, to some degree, from animal pollination. Many of these networks are freely available either as published studies {[}e.g., \protect\hyperlink{ref-carvalheiro2014}{43},\protect\hyperlink{ref-olesen2007}{44},\protect\hyperlink{ref-fortuna2010}{45}{]} or in online archives {[}e.g., Mangal \protect\hyperlink{ref-poisot2016}{46},The Web of Life \protect\hyperlink{ref-fortuna2010}{45}{]}. In total, our network dataset constituted 60 weighted (number of visits) and four unweighted (interaction presence-absence) networks, each sampled at a unique location and year, as well as eight meta-networks where interactions were pooled across several locations and multiple years.

@@

All species names, genera, families and orders were retrieved and standardized from the taxonomy data sources NCBI (\url{https://www.ncbi.nlm.nih.gov/taxonomy}) for plants and ITIS (\url{https://www.itis.gov/}) for pollinators, using the R package \emph{taxize} {[}\protect\hyperlink{ref-chamberlain2020}{47}{]}. We filled the `not found' searches manually using \url{http://www.theplantlist.org/} and \url{http://www.mobot.org/} for plants and \url{http://www.catalogueoflife.org/} for floral visitors.

@@

We selected a total of 19 quantitative and categorical functional traits that included reproductive traits and traits related with plant form and size (see table 1). Reproductive traits consisted of those related to floral and reproductive biology: (i) floral traits included those related to the size of floral organs (e.g., style length), floral display size (e.g., flower number) and floral rewards (e.g., pollen quantity), while (ii) reproductive biology traits represented the plant reproductive system (i.e., the breeding, mating and compatibility system). The remaining traits were related to plant size and form (i.e., plant height, lifespan and life form), which are commonly used to characterize the fast-slow continuum of plant trait variation {[}\protect\hyperlink{ref-salguero2016}{7}{]}. For each plant species, we undertook an extensive literature and online search for all traits across a wide range of resources (plant databases, online floras, books, journals and images). From a matrix comprising a total of 28,614 possible cells (considering all traits and plant species: 19 columns × 1,506 species), we were able to fill 23,354 cells (81.6\% of the dataset, see figure S2 for missing value information for each trait). An extended description of each trait and how it was obtained can be found in Appendix S1.

@@

We calculated the phylogenetic distance between all plant species considered using the function \emph{get\_tree} from the package \emph{rtrees} (\url{https://github.com/daijiang/rtrees}), which downloads phylogenetic distances from the extended R implementation of the Open Tree of Life {[}\protect\hyperlink{ref-smith2018}{48},\protect\hyperlink{ref-jin2019}{49}{]}.

@@

To avoid excluding species with trait missing values, we imputed trait missing values with the help of the function \emph{missForest} {[}\protect\hyperlink{ref-stekhoven2012}{50}{]} which allows imputation of datasets with continuous and categorical variables. We accounted for the phylogenetic distance among species on the imputation process by including the eigenvectors from principal component analysis of the phylogenetic distance (PCoA), which has been shown to improve the performance of \emph{missForest} {[}\protect\hyperlink{ref-penone2014}{51}{]}. To extract the eigenvectors, we used the function \emph{PVRdecomp} from the package \emph{PVR} {[}\protect\hyperlink{ref-santos2018}{52}{]} based on a previous conceptual framework that considers phylogenetic eigenvectors {[}\protect\hyperlink{ref-diniz-filho2012}{53}{]}. We conducted two different imputations, one for the full set of species (1,506 species: 5.79\% of missing values), and excluded nectar and pollen traits because of the high percentage of missing values (figure S2), and a second one for the subset of species with data for pollen per flower and microliters of nectar (755 species: 8.01\% of missing values). To corroborate that our imputation of missing values did not affect our results, we evaluated the reproductive spectrum (see section below) with and without imputed values and results for the two datasets were consistent (Fig S3 and Fig S4).

@@

We explored the association between the different quantitative plant traits using a phylogenetically informed Principal Component Analysis (pPCA). We did not include the quantitative variables of flower length and inflorescence width because they were highly and moderately correlated to flower width, respectively (Pearson's correlation = 0.72, \emph{P} \textless{} 0.01 and Pearson's correlation = 0.36, \emph{P} \textless{} 0.01), and thus we avoided overemphasizing flower size on the spectrum of trait variation. Prior to the analyses, we excluded outliers and standardized the data. Specifically, due to the high sensitivity of dimensionality reduction to outliers {[}\protect\hyperlink{ref-serneels2008}{54},\protect\hyperlink{ref-legendre2012}{55}{]}, we excluded values outside the 2.5th--97.5th percentile range, and thus our final dataset had 1,236 species. Then, we log transformed all continuous variables to reduce the influence of extreme values and then z-transformed (X= 0, SD=1) these variables so that they were within the same numerical range, as recommended for principal component analysis {[}\protect\hyperlink{ref-legendre2012}{55}{]}. Although qualitative traits were not included in the dimensionality reduction analysis, we also investigated the statistical association of the different qualitative traits with the main axes of trait variation using an ANOVA and Tukey test. We performed the pPCA using the function \emph{phyl.pca} from the package \emph{phytools} {[}\protect\hyperlink{ref-revell2012}{56}{]} with the method lambda (\(\lambda\)) that calculates the phylogenetic correlation between 0 (phylogenetic independence) and 1 (shared evolutionary history) and we implemented the mode ``covariance'' because values for each variables were on the same scale following transformation {[}\protect\hyperlink{ref-abdi2010}{57}{]}. Finally, we conducted an additional phylogenetic informed principal component analysis for the subset of species with pollen and nectar quantity data. For this, we included all quantitative traits included in the main pPCA, plus pollen grains and microlitres of nectar per flower.

@@

We calculated the phylogenetic signal of the different quantitative traits on the imputed dataset with the full set of species (N = 1,506) with the package \emph{phytools} {[}\protect\hyperlink{ref-revell2012}{56}{]} and used Pagel's \(\lambda\) as a measurement of the phylogenetic signal. However, for pollen and nectar traits, phylogenetic signal was calculated only on the subset of species that had quantitative information for these traits (N = 755).

@@

We used Bayesian modeling (see details below) to investigate how the different guilds of floral visitors interacted along the main axes of reproductive trait variation using qualitative and quantitative measures of plant-pollinator interactions. Specifically, interactions between plants and floral visitors in binary networks were defined as the presence or absence of floral visitation, while in weighted networks, interactions were measured as the the number of visits from each insect species to different plant species. Although floral visitors are not always pollinators and the number of visits does not consider the pollination efficiency of each species {[}\protect\hyperlink{ref-ballantyne2015}{58}{]}, the number of visits does provide valuable information about the relative contributions different floral visitors make to pollination {[}\protect\hyperlink{ref-vazquez2005}{59},\protect\hyperlink{ref-vazquez2012}{60}{]}. We divided floral visitors into six main taxonomical guilds that differ in life form, behaviour and thus likely play different ecological roles: (i) bees (Hymenoptera-Anthophila), (ii) non-bee-Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Diptera-Syrphidae), (iv) non-syrphid-Diptera (Diptera-non-Syrphidae), (v) Lepidoptera, and (vi) Coleoptera. Moreover, because bees had the greatest number of occurrences in networks (2,256 records) and the highest frequency of visits, we further broke this group down into the main bee families (Andrenidae, Apidae, Colletidae, Halictidae, and Megachilidae). In addition, we found that \emph{Apis mellifera} was the floral visitor with the largest proportion of occurrences in networks (7.55\% of the total). This is consistent with previous research showing that \emph{A. mellifera} was the most frequent floral visitor in a similar dataset of 80 plant-pollinator networks in natural ecosystems {[}\protect\hyperlink{ref-hung2018}{61}{]}. Hence, to control for the effect of \emph{A. mellifera} on observed visitation patterns of bees, we conducted an analogous analysis with presence-absence of the interaction and number of visits excluding \emph{A. mellifera}. We found that \emph{A. mellifera}, was partly driving some of the observed trends on PC1 (figure S5). However, we did not detect major differences on PC2 and PC3.

We used Bayesian generalized linear mixed models in the R package \emph{brms} {[}\protect\hyperlink{ref-burkner2017}{62}{]} to model the presence-absence of observed interactions and number of visits as a function of the main axes of plant trait variation and their interactions with floral visitor guilds (e.g., number of visits \textasciitilde{} PC1 x FGs + PC2 x FGs + PC3 x FGs). Because we were interested in possible differences in visitation patterns among floral visitor guilds to plants with different reproductive strategies, we included interactions between the main axes of trait variation (PC1, PC2 and PC3) and the different floral visitor guilds in the model. We added a nested random effect of networks within the study system to capture the variation in networks among studies and within networks. Moreover, we included the phylogenetic covariance matrix as a random factor due to the possible shared evolutionary histories of species and therefore lack of independence among them. We specified for the presence or absence of an interaction and number of visits with Bernoulli and a zero inflated negative binomial distributions, respectively. The models were run with 3,000 iterations with previous 1,000 warm up iterations and with non or very weakly informative priors from the brm function so the priors had negligible influence on the results {[}\protect\hyperlink{ref-burkner2017}{62}{]}. We set delta (\(\Delta\)) to 0.99 to avoid divergent transitions and visualized the posterior predictive checks with the function \emph{pp\_check} using the \emph{bayesplot} package {[}\protect\hyperlink{ref-gabry2019}{63}{]}.

@@

We also used Bayesian modeling to investigate how the main axes of trait variation determined different plant species' functional roles within pollination networks. For this, we selected complementary species-level network metrics, commonly applied in bipartite network studies {[}\protect\hyperlink{ref-dormann2008}{64}{]}, with clear ecological interpretations and are relevant for defining species' functional roles in networks. Specifically, the species-level metrics we included were: (i) sum of visits per plant species; (ii) normalized degree, calculated as the number of links per plant species divided by the total possible number of partners; and (iii) specialization (d') {[}\protect\hyperlink{ref-bluthgen2006}{65}{]}, which measures the deviation of an expected random choice of the available interaction partners and ranges between 0 (maximum generalization) and 1 (maximum specialization). Normalized degree and specialization were calculated with the \emph{specieslevel} function in the R package \emph{bipartite} {[}\protect\hyperlink{ref-dormann2008}{64}{]}. We modeled each plant species metric (sum of visits, normalized degree and plant specialization) as a function of the three main axes of trait variation (plant species level metric \textasciitilde{} PC1 + PC2 + PC3). For the different response variables in each model (i.e., each plant species level metric), we used different distribution families (zero inflated negative binomial for the sum of visits, weibull for normalized degree, and zero-one inflated beta for specialization). Finally, we used the same random factors, model settings and conducted the same posterior predictive checks for each model as detailed above in the `visitation patterns' section.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-barrett2002}{}}%
\CSLLeftMargin{1. }
\CSLRightInline{Barrett SCH. 2002 The evolution of plant sexual diversity. \emph{Nat. Rev. Genet.} \textbf{3}, 274--284. (doi:\href{https://doi.org/10.1038/nrg776}{10.1038/nrg776})}

>>>>

\setstretch{2}
\doublespacing
\normalsize

@@

Flowering plants display an astonishing diversity of reproductive structures (\protect\hyperlink{ref-barrett2002}{Barrett, 2002}; \protect\hyperlink{ref-schiestl2013}{Schiestl \& Johnson, 2013}) that shape their interactions with animal pollinators (\protect\hyperlink{ref-dellinger2020}{Dellinger, 2020}; \protect\hyperlink{ref-fenster2004}{Fenster et al., 2004}). The variety of reproductive strategies used by different plant species, from specialised reliance on a single pollinator, to full self-pollination, is incredibly rich. However, not all reproductive trait combinations in plants are possible due to evolutionary and ecological constraints (\protect\hyperlink{ref-agrawal2020}{Agrawal, 2020}; \protect\hyperlink{ref-stearns1989}{Stearns, 1989}). Despite recent advances in theoretical and empirical understanding of correlations between plant reproductive traits at large ecological scales (\protect\hyperlink{ref-friedman2020}{Friedman, 2020}; \protect\hyperlink{ref-paterno2020}{Paterno et al., 2020}; \protect\hyperlink{ref-roddy2021}{Roddy et al., 2021}; \protect\hyperlink{ref-salguero2016}{Salguero-Gómez et al., 2016}; \protect\hyperlink{ref-song2022}{Song et al., 2022}), studies typically focus on just a few traits, often in isolation. Thus, to develop a comprehensive understanding of the plant reproductive spectrum of trait variation, as done recently for root (\protect\hyperlink{ref-laughlin2021}{Laughlin et al., 2021}), leaf (\protect\hyperlink{ref-wright2004}{Wright et al., 2004}) and wood (\protect\hyperlink{ref-chave2009}{Chave et al., 2009}) related traits, we require a multi-trait perspective with broad geographical coverage. Importantly, characterization of the plant reproductive spectrum could reveal key reproductive strategies among different species (\protect\hyperlink{ref-agrawal2020}{Agrawal, 2020}) and improve knowledge of associations between specific reproductive traits and pollinators (\protect\hyperlink{ref-roddy2021}{Roddy et al., 2021}).

With the recent availability of large trait databases (e.g., TRY \protect\hyperlink{ref-kattge2011}{Kattge et al., 2011}; COMPADRE \protect\hyperlink{ref-salguero2015}{Salguero-Gómez et al., 2015}), there has been increased research on plant ecological strategies, which has produced identification of global patterns in plant form and function (\protect\hyperlink{ref-bruelheide2018}{Bruelheide et al., 2018}; \protect\hyperlink{ref-carmona2021}{Carmona et al., 2021}; \protect\hyperlink{ref-diaz2016}{Díaz et al., 2016}; \protect\hyperlink{ref-salguero2016}{Salguero-Gómez et al., 2016}). However, studies that look at multiple traits tend to overlook plant reproductive traits (\protect\hyperlink{ref-evojtko2020}{E-Vojtkó et al., 2020}; \protect\hyperlink{ref-roddy2021}{Roddy et al., 2021}; \protect\hyperlink{ref-ruger2018}{Rüger et al., 2018}), and focus on trait correlations in other plant organs such as morphological and physiological leaf traits (\protect\hyperlink{ref-donovan2011}{Donovan et al., 2011}; \protect\hyperlink{ref-osnas2013}{Osnas et al., 2013}; \protect\hyperlink{ref-shipley2006}{Shipley et al., 2006}; \protect\hyperlink{ref-wright2004}{Wright et al., 2004}). Despite the lack of a holistic understanding that depicts reproductive trait covariation patterns, there are widely recognized associations between plant reproductive trait pairs such as the negative correlation between flower size and flower number (\protect\hyperlink{ref-vojtko2022}{E-Vojtkó et al., 2022}; \protect\hyperlink{ref-kettle2011}{Kettle et al., 2011}; \protect\hyperlink{ref-sargent2007}{Sargent et al., 2007}); the positive association between flower size and outcrossing rate (\protect\hyperlink{ref-goodwillie2010}{Goodwillie et al., 2010}); or the positive correlation between outcrossing rate and lifespan (\protect\hyperlink{ref-barrett2003}{Barrett, 2003}; \protect\hyperlink{ref-moeller2017}{Moeller et al., 2017}; \protect\hyperlink{ref-munoz2016}{Munoz et al., 2016})). Although identification of these trait correlations has allowed progress toward a conceptual framework that integrates different floral trait relationships (\protect\hyperlink{ref-roddy2021}{Roddy et al., 2021}), we still lack empirical understanding of how these associations among floral and other reproductive traits (e.g., reproductive biology) shape plant life history strategies.

Trait-based approaches (\protect\hyperlink{ref-fenster2004}{Fenster et al., 2004}; \protect\hyperlink{ref-rosas2014}{Rosas-Guerrero et al., 2014}) and trait-matching analyses (\protect\hyperlink{ref-bartomeus2016}{Bartomeus et al., 2016}; \protect\hyperlink{ref-stang2009}{Stang et al., 2009}) are powerful for exploring the underlying mechanisms driving plant-pollinator interactions. For example, plant traits can define species' network roles (e.g., specialists vs generalists \protect\hyperlink{ref-tur2013}{Tur et al., 2013}) and plant species that occupy reproductive trait space extremes are likely to be more specialised and display tighter trait-matching with pollinators (\protect\hyperlink{ref-coux2016}{Coux et al., 2016}; \protect\hyperlink{ref-junker2013}{Junker et al., 2013}). Indeed, the degree of morphological matching between plants and floral visitors often determines whether or not interactions occur, and consequently influences the broader network structure (\protect\hyperlink{ref-ibanez2012}{Ibanez, 2012}; \protect\hyperlink{ref-stang2009}{Stang et al., 2009}). However, it is still unclear how specific reproductive traits (e.g., mating or compatibility system) influence plant-pollinator associations (\protect\hyperlink{ref-devaux2014}{Devaux et al., 2014}; \protect\hyperlink{ref-sargent2008}{Sargent \& Ackerly, 2008}; \protect\hyperlink{ref-tur2013}{Tur et al., 2013}). Indeed, both morphology and mating system can determine a species' functional role within pollination networks, and the combination of several traits increases power for predicting network structure (\protect\hyperlink{ref-eklof2013}{Eklöf et al., 2013}). Because the species position within the multidimensional trait space can result in different interaction patterns or adaptations to pollinators (see \protect\hyperlink{ref-dehling2016}{Dehling et al., 2016} for a purely morphological approach), the use of the reproductive trait space to explore plant-pollinator interactions could help to identify species functional roles and associations of traits with specific pollinators (i.e., pollination syndromes).

@@

We selected 28 studies from 18 different countries that constituted a total of 64 plant-pollinator networks (see table S1 and figure S1). These studies recorded plant-pollinator interactions in natural systems and were selected so that we had broad geographical representation across different ecological communities. Although these studies differ in sampling effort and methodology, all studies provided information about plant-pollinator interactions (weighted and non-weighted), which we used to build a database of plant species that are likely to benefit, to some degree, from animal pollination. Many of these networks are freely available either as published studies (e.g., \protect\hyperlink{ref-carvalheiro2014}{Carvalheiro et al., 2014}; \protect\hyperlink{ref-fortuna2010}{Fortuna et al., 2010}; \protect\hyperlink{ref-olesen2007}{Olesen et al., 2007}) or in online archives (e.g., Mangal \protect\hyperlink{ref-poisot2016}{Poisot et al., 2016}; The Web of Life \protect\hyperlink{ref-fortuna2010}{Fortuna et al., 2010}). In total, our network dataset constituted 60 weighted (number of visits) and four unweighted (interaction presence-absence) networks, each sampled at a unique location and year, as well as eight meta-networks where interactions were pooled across several locations and multiple years.

@@

All species names, genera, families and orders were retrieved and standardised from the taxonomy data sources NCBI (\url{https://www.ncbi.nlm.nih.gov/taxonomy}) for plants and ITIS (\url{https://www.itis.gov/}) for pollinators, using the R package \emph{taxize} (\protect\hyperlink{ref-chamberlain2020}{Chamberlain et al., 2020}). We filled the `not found' searches manually using \url{http://www.theplantlist.org/} and \url{http://www.mobot.org/} for plants and \url{http://www.catalogueoflife.org/} for floral visitors.

@@

We selected a total of 19 quantitative and categorical functional traits that included reproductive traits and traits related with plant form and size (see table 1). Reproductive traits consisted of those related to floral and reproductive biology: (i) floral traits included those related to the size of floral organs (e.g., style length), floral display size (e.g., flower number) and floral rewards (e.g., pollen quantity), while (ii) reproductive biology traits represented the plant reproductive system (i.e., the breeding, mating and compatibility system). The remaining traits were related to plant size and form (i.e., plant height, lifespan and life form), which are commonly used to characterise the fast-slow continuum of plant trait variation (\protect\hyperlink{ref-salguero2016}{Salguero-Gómez et al., 2016}). For each plant species, we undertook an extensive literature and online search for all traits across a wide range of resources (plant databases, online floras, books, journals and images). From a matrix comprising a total of 28,614 possible cells (considering all traits and plant species: 19 columns × 1,506 species), we were able to fill 23,354 cells (81.6\% of the dataset, see figure S2 for missing value information for each trait). An extended description of each trait and how it was obtained can be found in Appendix S1.

@@

We calculated the phylogenetic distance between all plant species considered using the function \emph{get\_tree} from the package \emph{rtrees} (\url{https://github.com/daijiang/rtrees}), which downloads phylogenetic distances from the extended R implementation of the Open Tree of Life (\protect\hyperlink{ref-jin2019}{Jin \& Qian, 2019}; \protect\hyperlink{ref-smith2018}{Smith \& Brown, 2018}).

@@

To avoid excluding species with trait missing values, we imputed trait missing values with the help of the function \emph{missForest} (\protect\hyperlink{ref-stekhoven2012}{Stekhoven \& Bühlmann, 2012}) which allows imputation of datasets with continuous and categorical variables. We accounted for the phylogenetic distance among species on the imputation process by including the eigenvectors from principal component analysis of the phylogenetic distance (PCoA), which has been shown to improve the performance of \emph{missForest} (\protect\hyperlink{ref-penone2014}{Penone et al., 2014}). To extract the eigenvectors, we used the function \emph{PVRdecomp} from the package \emph{PVR} (\protect\hyperlink{ref-santos2018}{Chamberlain et al., 2018}) based on a previous conceptual framework that considers phylogenetic eigenvectors (\protect\hyperlink{ref-diniz-filho2012}{Diniz-Filho et al., 2012}). We conducted two different imputations, one for the full set of species (1,506 species: 5.79\% of missing values), and excluded nectar and pollen traits because of the high percentage of missing values (figure S2), and a second one for the subset of species with data for pollen per flower and microliters of nectar (755 species: 8.01\% of missing values). To corroborate that our imputation of missing values did not affect our results, we evaluated the reproductive spectrum (see section below) with and without imputed values and results for the two datasets were consistent (Fig S3 and Fig S4).

@@

We explored the association between the different quantitative plant traits using a phylogenetically informed Principal Component Analysis (pPCA). We did not include the quantitative variables of flower length and inflorescence width because they were highly and moderately correlated to flower width, respectively (Pearson's correlation = 0.72, \emph{P} \textless{} 0.01 and Pearson's correlation = 0.36, \emph{P} \textless{} 0.01), and thus we avoided overemphasise flower size on the spectrum of trait variation. Prior to the analyses, we excluded outliers and standardised the data. Specifically, due to the high sensitivity of dimensionality reduction to outliers (\protect\hyperlink{ref-legendre2012}{Legendre \& Legendre, 2012}; \protect\hyperlink{ref-serneels2008}{Serneels \& Verdonck, 2008}), we excluded values outside the 2.5th--97.5th percentile range, and thus our final dataset had 1,236 species. Then, we log transformed all continuous variables to reduce the influence of extreme values and then z-transformed (X= 0, SD=1) these variables so that they were within the same numerical range, as recommended for principal component analysis (\protect\hyperlink{ref-legendre2012}{Legendre \& Legendre, 2012}). Although qualitative traits were not included in the dimensionality reduction analysis, we also investigated the statistical association of the different qualitative traits with the main axes of trait variation using an ANOVA and Tukey test. We performed the pPCA using the function \emph{phyl.pca} from the package \emph{phytools} (\protect\hyperlink{ref-revell2012}{Revell, 2012}) with the method lambda (\(\lambda\)) that calculates the phylogenetic correlation between 0 (phylogenetic independence) and 1 (shared evolutionary history) and we implemented the mode ``covariance'' because values for each variables were on the same scale following transformation (\protect\hyperlink{ref-abdi2010}{Abdi \& Williams, 2010}). Finally, we conducted an additional phylogenetic informed principal component analysis for the subset of species with pollen and nectar quantity data. For this, we included all quantitative traits included in the main pPCA, plus pollen grains and microlitres of nectar per flower.

@@

We calculated the phylogenetic signal of the different quantitative traits on the imputed dataset with the full set of species (N = 1,506) with the package \emph{phytools} (\protect\hyperlink{ref-revell2012}{Revell, 2012}) and used Pagel's \(\lambda\) as a measurement of the phylogenetic signal. However, for pollen and nectar traits, phylogenetic signal was calculated only on the subset of species that had quantitative information for these traits (N = 755).

@@

We used Bayesian modelling (see details below) to investigate how the different guilds of floral visitors interacted along the main axes of reproductive trait variation using qualitative and quantitative measures of plant-pollinator interactions. Specifically, interactions between plants and floral visitors in binary networks were defined as the presence or absence of floral visitation, while in weighted networks, interactions were measured as the the number of visits from each insect species to different plant species. Although floral visitors are not always pollinators and the number of visits does not consider the pollination efficiency of each species (\protect\hyperlink{ref-ballantyne2015}{Ballantyne et al., 2015}), the number of visits does provide valuable information about the relative contributions different floral visitors make to pollination (\protect\hyperlink{ref-vazquez2005}{Vázquez et al., 2005}, \protect\hyperlink{ref-vazquez2012}{2012}). We divided floral visitors into six main taxonomical guilds that differ in life form, behaviour and thus likely play different ecological roles: (i) bees (Hymenoptera-Anthophila), (ii) non-bee-Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Diptera-Syrphidae), (iv) non-syrphid-Diptera (Diptera-non-Syrphidae), (v) Lepidoptera, and (vi) Coleoptera. Moreover, because bees had the greatest number of occurrences in networks (2,256 records) and the highest frequency of visits, we further broke this group down into the main bee families (Andrenidae, Apidae, Colletidae, Halictidae, and Megachilidae). In addition, we found that \emph{Apis mellifera} was the floral visitor with the largest proportion of occurrences in networks (7.55\% of the total). This is consistent with previous research showing that \emph{A. mellifera} was the most frequent floral visitor in a similar dataset of 80 plant-pollinator networks in natural ecosystems (\protect\hyperlink{ref-hung2018}{Hung et al., 2018}). Hence, to control for the effect of \emph{A. mellifera} on observed visitation patterns of bees, we conducted an analogous analysis with presence-absence of the interaction and number of visits excluding \emph{A. mellifera}. We found that \emph{A. mellifera}, was partly driving some of the observed trends on PC1 (figure S5). However, we did not detect major differences on PC2 and PC3.

We used Bayesian generalised linear mixed models in the R package \emph{brms} (\protect\hyperlink{ref-burkner2017}{Bürkner, 2017}) to model the presence-absence of observed interactions and number of visits as a function of the main axes of plant trait variation and their interactions with floral visitor guilds (e.g., number of visits \textasciitilde{} PC1 x FGs + PC2 x FGs + PC3 x FGs). Because we were interested in possible differences in visitation patterns among floral visitor guilds to plants with different reproductive strategies, we included interactions between the main axes of trait variation (PC1, PC2 and PC3) and the different floral visitor guilds in the model. We added a nested random effect of networks within the study system to capture the variation in networks among studies and within networks. Moreover, we included the phylogenetic covariance matrix as a random factor due to the possible shared evolutionary histories of species and therefore lack of independence among them. We specified for the presence or absence of an interaction and number of visits with Bernoulli and a zero inflated negative binomial distributions, respectively. The models were run with 3,000 iterations with previous 1,000 warm up iterations and with non or very weakly informative priors from the brm function so the priors had negligible influence on the results (\protect\hyperlink{ref-burkner2017}{Bürkner, 2017}). We set delta (\(\Delta\)) to 0.99 to avoid divergent transitions and visualized the posterior predictive checks with the function \emph{pp\_check} using the \emph{bayesplot} package (\protect\hyperlink{ref-gabry2019}{Gabry et al., 2019}).

@@

We also used Bayesian modelling to investigate how the main axes of trait variation determined different plant species' functional roles within pollination networks. For this, we selected complementary species-level network metrics, commonly applied in bipartite network studies (\protect\hyperlink{ref-dormann2008}{Dormann et al., 2008}), with clear ecological interpretations and are relevant for defining species' functional roles in networks. Specifically, the species-level metrics we included were: (i) sum of visits per plant species; (ii) normalised degree, calculated as the number of links per plant species divided by the total possible number of partners; and (iii) specialisation (d') (\protect\hyperlink{ref-bluthgen2006}{Blüthgen et al., 2006}), which measures the deviation of an expected random choice of the available interaction partners and ranges between 0 (maximum generalisation) and 1 (maximum specialisation). Normalised degree and specialisation were calculated with the \emph{specieslevel} function in the R package \emph{bipartite} (\protect\hyperlink{ref-dormann2008}{Dormann et al., 2008}). We modelled each plant species metric (sum of visits, normalised degree and plant specialisation) as a function of the three main axes of trait variation (plant species level metric \textasciitilde{} PC1 + PC2 + PC3). For the different response variables in each model (i.e., each plant species level metric), we used different distribution families (zero inflated negative binomial for the sum of visits, weibull for normalised degree, and zero-one inflated beta for specialisation). Finally, we used the same random factors, model settings and conducted the same posterior predictive checks for each model as detailed above in the `visitation patterns' section.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-abdi2010}{}}%
Abdi, H., \& Williams, L. J. (2010). Principal component analysis. \emph{WIREs Comp. Stats.}, \emph{2}(4), 433--459. \url{https://doi.org/10.1002/wics.101}
